### 2015-02-27 icfp15-servo/abstract.tex
Add a pic. Clean up the abstract a bit

%!TEX root = paper.tex
%
\begin{abstract}
Servo is a project started at Mozilla Research to build a new web browser
engine that takes advantage of both modern parallel hardware and advances in language
technology that prevent many of the common security errors that plague all of the other
modern browsers.
The new language, Rust, enabled us to take on this project in a type-centric way instead of
the traditional C++ that relies on coding conventions, imprecise static analyses, and
fuzzers.
While many languages before have offered some combination of memory safety, parallelism,
and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
with the control over both memory and generated code that is required by systems programmers.
In this paper, \emph{systems} \emph{software} refers to programs that have some combination
of soft realtime requirements and memory, CPU, and power usage constraints that require the
programmer to be able to predict these features from their source code --- without being a
compiler developer.

Servo is currently over 300k lines of code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.
In this work, we show how a language with an affine type system and many syntactic
features from the ML family of languages can be successfully used to build state-of-the-art
systems software.
We also outline several pitfalls encountered along the way and describe some potential
areas for future research.
\end{abstract}

>>>>

%!TEX root = paper.tex
%
\begin{abstract}
All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Webkit ---
have a core rendering engine written in \Cplusplus.
This language choice was made because it affords the systems programmer complete control
of the underlying hardware features and memory in use plus provides a transparent
compilation model.

Servo is a project started at Mozilla Research to build a new web browser engine that
preserves the capabilities of these other browser engines but also both takes more
advantage of the recent trends in parallel hardware and is more memory-safe.
We use a new language, Rust, that provides us a similar level of control of the underlying
system but which builds on many concepts familiar to the functional programming community
to form a novelty - a useful, safe systems programming language.

In this paper, we show how a language with an affine type system and many syntactic
features familiar to functional language programmers can be successfully used to build
state-of-the-art systems software.
We also outline several pitfalls encountered along the way and describe some potential
areas for future research.
\end{abstract}

### 2015-02-27 icfp15-servo/browsers.tex
Add a pic. Clean up the abstract a bit

Modern web browsers do not just load static pages, but can also handle pages that have similar
complexity to native applications.
From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
based on the Unreal Engine\footnote{\url{https://blog.mozilla.org/blog/2015/02/24/unreal-engine-4-7-binary-release-includes-html5-export-3/}},
modern browsers have the ability to
While Servo is designed to handle applications of that complexity, in this section we will
focus on the subset of the browser design that supports loading and displaying a single page.

>>>>

Modern web browsers do not just load static pages, but can also handle pages that have similar
complexity to native applications.
From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
based on the Unreal Engine\footnote{\url{https://blog.mozilla.org/blog/2015/02/24/unreal-engine-4-7-binary-release-includes-html5-export-3/}},
modern browsers have the ability to handle much more than simple static pages.
\figref{fig:browser} shows the steps in processing a site.
While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[scale=0.7]{pics/browser}
  \end{center}%
  \caption{Processing stages and intermediate representations in a browser engine.}
  \label{fig:browser}
\end{figure*}%

### 2015-02-27 icfp15-servo/intro.tex
Add a pic. Clean up the abstract a bit

\section{Introduction}
\label{sec:intro}
Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
was started in February of 2013.

Modern web browser engines have a similar complexity level to that of operating systems.
These engines are expected to execute arbitrary user programs within a safe sandbox, scale 
across a wide variety of hardware, manage access to shared system resources, and further 
do all of that while providing good performance and correct behavior
against sometimes-incomplete standards.
With the expectation of executing well on mobile devices and parallel hardware, the 
fundamental design assumptions have significantly changed since all of the modern
web engines were originally designed.

>>>>

\section{Introduction}
\label{sec:intro}
At the heart of a modern web browser is the browser engine, which is the code responsible
for loading, processing, evaluating, and rendering web content.
There are three current browser engine families:
\begin{enumerate}
\item Trident/Spartan, the engine in Internet Explorer
\item Webkit/Blink, the engine in Safari, Chrome, and Opera
\item Gecko, the engine in Firefox
\end{enumerate}
All of these engines have at their core many millions of lines of \Cplusplus code.
While the use of \Cplusplus has enabled all of these browsers to achieve excellent sequential
performance on a single web page, on mobile devices with lower processor speed but many
more processors, these browsers do not provide the same level of interactivity that they
do on desktop processors~\cite{parallelizing-web-pages,ZOOMM}.
Further, in an informal inspection of the critical security bugs in Gecko, we determined that
roughly 50\% of the bugs are use after free, out of range access, or related to integer
overflow.
The other 50\% is split between errors in tracing values from the JavaScript (JS) heap in the
\Cplusplus code and errors related to Just-in-time (JIT) code, which we will not discuss
further in this work.

Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
was started in February of 2013.
The goal of this project is to produce a browser that enables new applications to be authored
against the web platform that run with more safety, better performance, and better power usage
than in current browsers.
Modern web browser engines have a similar complexity level to that of operating systems.
These engines are expected to execute arbitrary user programs within a safe sandbox, scale 
across a wide variety of hardware, manage access to shared system resources, and further 
do all of that while providing good performance and correct behavior
against sometimes-incomplete standards.
With the expectation of executing well on mobile devices and parallel hardware, the 
fundamental design assumptions have significantly changed since all of the modern
web engines were originally designed.

In this paper, \emph{systems} \emph{software} refers to programs that have some combination
of soft realtime requirements and memory, CPU, and power usage constraints that require the
programmer to be able to predict these features from their source code --- without being a
compiler developer.

enabled us to take on this project in a type-centric way instead of
the traditional C++ that relies on coding conventions, imprecise static analyses, and
fuzzers.
While many languages before have offered some combination of memory safety, parallelism,
and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
with the control over both memory and generated code that is required by systems programmers.
Servo is currently over 300k lines of code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.

### 2015-02-27 icfp15-servo/rust.tex
Add a pic. Clean up the abstract a bit

\section{Open}
\label{sec:open}

>>>>

\section{Rust}
\label{sec:rust}

TODO: basic overview

\subsection{Syntactic features}

Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
well as pattern matching.
These types and associated language features provided two large benefits to Servo over traditional browsers
written in \Cplusplus{}.
First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
files.
More importantly, this style of programming that uses pattern matching over concrete types instead of
virtual function calls on a class hierarchy provides a non-trivial performance gain.
Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
incur indirect function call costs.
All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
specifier wherever possible or specialize the code in some other way to avoid this cost.

Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
programs from languages such as \Cplusplus.
This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
with their attempts to build a safe language that preserved most C idioms.

\subsection{Language interoperability}
Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially. This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.

Would like to have full C++ inlining.

\subsection{Compilation strategy}
Monomorphization, as in MLton, is a great way of providing predictable output code to developers.

Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.

\subsection{Memory management}
Rust has an ownership model.
Unlike ML Kit (unbounded regions) and Singularity (ownership bleeding into interfaces), Rust has both statically scoped regions and a notion of borrowing.
Statically scoped regions

Would like to be able to plug in our own GC and stack scanning.
